---
title: "Zarr Operations Cookbook"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Zarr Operations Cookbook}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
```

This vignette covers common zarr array operations: persistent storage,
compression, resizing, filters, and advanced indexing.

```{r}
library(pizzarr)
```

## Persistent arrays

Create an array on disk, close the session, and reopen it later.

```{r}
path <- file.path(tempdir(), "example.zarr")

# Create a persistent array backed by a DirectoryStore
z <- zarr_open_array(
  store = path, mode = "w",
  shape = c(5, 10), chunks = c(5, 5), dtype = "<f4"
)

# Write data
z$set_item("...", array(1:50, dim = c(5, 10)))

z$get_shape()
```

Reopen the same path in read mode:

```{r}
z2 <- zarr_open_array(store = path, mode = "r")

z2$get_shape()

z2$get_item("...")$data
```

For quick save/load of an existing array:

```{r}
save_path <- file.path(tempdir(), "saved.zarr")

# Save an R array directly
zarr_save_array(save_path, zarr_create_array(
  data = volcano, shape = dim(volcano), dtype = "<f8"
))

# Reopen
z3 <- zarr_open_array(save_path, mode = "r")

all.equal(z3$as.array(), volcano)
```

## Compression

By default, pizzarr uses Zstandard compression. You can choose a different
compressor when creating an array.

### Zstandard (default)

```{r}
z_zstd <- zarr_create(
  shape = c(100, 100), dtype = "<f4",
  compressor = ZstdCodec$new(level = 3)
)

z_zstd$get_compressor()$get_config()
```

### Gzip

```{r}
z_gzip <- zarr_create(
  shape = c(100, 100), dtype = "<f4",
  compressor = GzipCodec$new(level = 5)
)

z_gzip$get_compressor()$get_config()
```

### Blosc (with algorithm selection)

```{r}
z_blosc <- zarr_create(
  shape = c(100, 100), dtype = "<f4",
  compressor = BloscCodec$new(cname = "lz4", clevel = 5, shuffle = TRUE)
)

z_blosc$get_compressor()$get_config()
```

### No compression

```{r}
z_none <- zarr_create(
  shape = c(100, 100), dtype = "<f4",
  compressor = NA
)

is.na(z_none$get_compressor())
```

## Resizing arrays

Arrays can be resized after creation. Data in the overlapping region is
preserved; new regions are filled with the fill value.

```{r}
z <- zarr_create(
  shape = c(5, 10), chunks = c(5, 5),
  dtype = "<i4", fill_value = 0L,
  compressor = "default"
)

z$set_item("...", array(1:50, dim = c(5, 10)))

z$get_shape()

# Grow the array
z$resize(10, 20)

z$get_shape()

# Original data is preserved in the top-left corner
z[1:5, 1:10]$data

# New region is filled with fill_value
z[6:10, 1:5]$data
```

Shrinking removes chunks that fall outside the new shape:

```{r}
z$resize(3, 4)

z$get_shape()

z$get_item("...")$data
```

## Appending data

Use `append()` to grow an array along an axis, adding new data at the end.
This is equivalent to zarr-python's `z.append(data, axis=0)`, but uses
R's 1-based axis indexing (axis 1 = first dimension).

```{r}
z <- zarr_create(
  shape = c(3, 4), chunks = c(3, 4),
  dtype = "<i4", fill_value = 0L
)

z$set_item("...", array(1:12, dim = c(3, 4)))

z$as.array()
```

Append new rows (axis 1, the default):

```{r}
new_rows <- array(13:20, dim = c(2, 4))

z$append(new_rows)

z$get_shape()

z$as.array()
```

Append new columns (axis 2):

```{r}
new_cols <- array(21:30, dim = c(5, 2))

z$append(new_cols, axis = 2)

z$get_shape()

z$as.array()
```

## Filters

Filters transform chunk data before compression. They are codec instances
passed as a list to the `filters` parameter. A common use case is
variable-length UTF-8 string arrays, which require `VLenUtf8Codec` as a filter.

```{r}
words <- c("alpha", "bravo", "charlie", "delta")

z_str <- zarr_create_array(
  data = array(words, dim = length(words)),
  shape = length(words), dtype = "|O",
  object_codec = VLenUtf8Codec$new()
)

z_str$get_item("...")$data

z_str$get_filters()
```

## Advanced indexing

Beyond basic slicing with `slice()` or `[`, pizzarr supports orthogonal
indexing for independent selection along each dimension.

### Setup

```{r}
z <- zarr_create_array(
  data = matrix(1:30, nrow = 5, ncol = 6),
  shape = c(5, 6), dtype = "<i4"
)

z$as.array()
```

### Basic slicing with `[`

The bracket operator uses orthogonal indexing internally:

```{r}
# Select rows 1-3, columns 2-4
z[1:3, 2:4]$data
```

### Orthogonal selection with integer arrays

Select specific rows and columns independently. Note that
`get_orthogonal_selection` uses zero-based indices (like zarr-python),
while the `[` operator uses R's one-based indexing:

```{r}
z$get_orthogonal_selection(list(c(0L, 2L, 4L), zb_slice(0, 6)))$data
```

### Boolean (mask) dimension indexing

Select dimensions using logical vectors:

```{r}
row_mask <- c(TRUE, FALSE, TRUE, FALSE, TRUE)

z$get_orthogonal_selection(list(row_mask, zb_slice(0, 6)))$data
```

### Using the OIndex object

The `$get_oindex()` accessor provides the same orthogonal indexing:

```{r}
oi <- z$get_oindex()

oi$get_item(list(c(0L, 4L), c(1L, 3L, 5L)))$data
```

### Slicing with step

Select every other row, every third column using `seq()` in bracket notation:

```{r}
z[seq(1, 5, 2), seq(1, 6, 3)]$data
```

### Ellipsis and colon shorthand

`"..."` selects all remaining dimensions; `":"` selects all along one dimension.
These work with `get_item()`:

```{r}
# All rows, column 1
z$get_item(list(":", 1))$data

# Row 1, all columns
z$get_item(list(1, "..."))$data
```

```{r, include = FALSE}
unlink(file.path(tempdir(), "example.zarr"), recursive = TRUE)
unlink(file.path(tempdir(), "saved.zarr"), recursive = TRUE)
```
